Phylogenetics and the human microbiome.

### 2014-04-14 reviews/response.txt

> 7-23 delete though

Nice catch!

> The section "The Human Microbiome" is quite detailed (discussing mouse
> to human transplants wrt obesity, etc.). Given the intense media
> reporting of the work coming from this area, SB readers may not need
> to be primed of the importance of this field. If this were shortened,
> there could be room to expand some of the technical bits that SB
> readers will want that follow. The last part of the section  - about
> sticking to eubacteria and a non-molecular biology approach do seem
> germane.

The editors have stated that they are not concerned with this becoming
too long, and I've been surprised that what seems like an absurdly big
amount of media hype to me doesn't reach everyone. So I'm leaving it in
unless the editors say otherwise.

@@

> 8-12. these data (I guess we will eventually decide that data is
> singular - up to the editors)

I have fixed all these.

> 8-37++ Whereas Woese ++ This is a bit wordy and unnecessary - we know
> that Woese didn't have PCR

Er, perhaps this is a generational thing? I had no idea how they used to
infer molecular sequences before Pace's phyloseminar.

@@

> 10-30 Which "databases" ? - this has been written so far as if it were
> being applied to samples, not databases. I am not clear what the unit
> of observation is here.

I have clarified with a slightly expanded explanation, thanks.

@@

> 12-37++. Is the MEGAN approach compositional or phylogenetic? I
> believe it is meant to be a hybrid approach (BLAST being
> compositional, but the taxonomic labeling considered phylogenetic),
> but this could be clarified.

I've added in a clarification, thanks.

> 13-29. It is not clear what heuristics are required for a cut-off
> method. The author says that he is going to describe the OTU inference
> industry, but the paragraph is really just a list of acronyms, and
> then refers obliquely to a "contingency table" of OTU observations.
> So, for example, what is a notion of goodness of clustering that the
> heuristics are trying to achieve? What are the sorts of species
> divisions found in taxonomies? This is an area where more would be
> helpful - and where a shorter introduction might offer space.

These are excellent questions, and the landscape is a little complex. I
have expanded this section to explain what it is that these methods
would like to achieve.

### 2014-04-14 sbreview.tex
Done with Mooers review!

It is possible to use metagenomic data as an expanded set of marker genes.
That is, one can use 16S reads that appear in the metagenome as well as reads from other ``core" genes that are expected to follow the same evolutionary path and are present in a large proportion of micro-organisms.
This is proven to be a useful strategy \citep{von2007quantitative,wu2008amphora,stark2010mltreemap,kembel2011phylogenetic}, however, because of the variability of gene repertoire in microbes, this core gene set may be relatively small.
Indeed, even the largest collection of genes in these databases only recruits around 1 percent of a metagenome.
At least some portion of the rest of the data can be taxonomically classified \citep[methods reviewed by][]{mande2012classification}; \citet{treangen2013metamos} report speedups and much higher accuracy when reads are assembled before they are classified.

@@

Several groups are continually revising taxonomy using the results of phylogenetic tree inference.
These attempts are less ambitious than the PhyloCode project \citep[to develop a taxonomic scheme expressed directly in terms of a phylogeny; see][]{forey2001phylocode}, and simply work to revise the hierarchical structure of the taxonomy while for the most part leaving taxonomic names fixed.
% to preserve information associated with taxonomic classifications.
Bergey's Manual of Systematic Bacteriology has officially adopted 16S as the basis for their taxonomy, although the actual revision process appears opaque \citep{kreig1984bergey}.
The GreenGenes group \citep{desantis2006greengenes} has been very active in updating their taxonomy according to 16S, first with their GRUNT tool \citep{dalevi2007automated} and more recently with their tax2tree tool \citep{mcdonald2011improved}.
Tax2tree uses a heuristic algorithm to optimize the $F$-measure of precision and recall for its taxonomic assignments.
Interestingly, tax2tree allows for polyphyletic taxonomic groups as allowance for either phylogenetic error, lack of resolution of the 16S gene or taxonomic groups with no evolutionary basis; the method does not attempt to signal the cause of such polyphyletic groups.
\citet{matsen2012reconciling} developed algorithms to quantify discordance between phylogeny and taxonomy based on a coloring problem previously described in the computer science literature \citep{moran2008convex}.
Although it is wonderful that several groups are actively working on taxonomic revision, it can be frustrating to have multiple different taxonomies with no easy way to translate between them or to the taxonomic names provided in the NCBI sequence database.

An obvious application of phylogenetics is to perform taxonomic classification, as the taxonomy is at least in part defined by phylogeny.
However, comparisons of taxonomic classification programs \citep{liu2008accurate,bazinet2012comparative} have indicated that current implementations of phylogenetic methods do not perform as well as simple classifiers based on $k$-mer composition \citep{wang2007naive,rosen2008metagenome}.
\citet{srinivasan2012bacterial} find that phylogenetic methods to do taxonomic classification can outperform composition-based techniques at least for certain taxonomic groups.
% \citep{werner2011impact} Impact of training sets on classification of high-throughput bacterial 16s rRNA gene surveys
Some authors report that a combination of composition-based and homology-based classifiers work best \citep{brady2009phymm,parks2011classifying}.
The MEGAN program \citep{huson2007megan,huson2011integrative} BLASTs an unknown sequence onto a database of sequences with taxonomic labels and assigns the sequence the lowest (i.e.\ narrowest) taxonomic group shared by all of the high-quality hits.
\citet{munch2008statistical,munch2008fast} infer taxonomic assignment by automatically retrieving sequences equipped with taxonomic information and building a tree on them along with an unknown sequence.
\citet{segata2012metagenomic} propose a clever approach to inferring organisms present in a metagenomic sample by compiling a database of clade-specific genes, then classifying a given read as being from the only clade that has the corresponding gene.
They show that this has good sensitivity and specificity, however, this method can only be used to identify the presence of organisms whose genome has been sequenced.

>>>>

It is possible to use metagenomic data as an expanded set of marker genes.
That is, one can use 16S reads that appear in the metagenome as well as reads from other ``core" genes that are expected to follow the same evolutionary path and are present in a large proportion of micro-organisms.
This is proven to be a useful strategy, and several groups have built databases of core gene families and provide programs and/or web tools to subset a metagenome to that core set of genes \citep{von2007quantitative,wu2008amphora,stark2010mltreemap,kembel2011phylogenetic,Darling2014-ux}.
However, because of the variability of gene repertoire in microbes, this core gene set may be relatively small: even the largest collection of genes in these databases only recruits around 1 percent of a metagenome.
At least some portion of the rest of the other approximately 99 percent of the metagenome can be taxonomically classified using methods reviewed by \citep{mande2012classification}.
\citet{treangen2013metamos} report shorter run times and much higher accuracy (for taxonomic classification using reads simulated from full genomes) for such metagenomic classification when reads are assembled before they are classified.

@@

Several groups are continually revising taxonomy using the results of phylogenetic tree inference.
These attempts are less ambitious than the PhyloCode project \citep[to develop a taxonomic scheme expressed directly in terms of a phylogeny; see][]{forey2001phylocode}, and simply work to revise the hierarchical structure of the taxonomy while for the most part leaving taxonomic names fixed.
% to preserve information associated with taxonomic classifications.
Bergey's Manual of Systematic Bacteriology has officially adopted 16S as the basis for their taxonomy, although the actual revision process appears opaque \citep{kreig1984bergey}.
The GreenGenes group \citep{desantis2006greengenes} has been very active in updating their taxonomy according to 16S, first with their GRUNT tool \citep{dalevi2007automated} and more recently with their tax2tree tool \citep{mcdonald2011improved}.
Tax2tree uses a heuristic algorithm to optimize the $F$-measure (the harmonic mean of precision and recall) for its taxonomic assignments.
Interestingly, tax2tree allows for polyphyletic taxonomic groups as allowance for either phylogenetic error, lack of resolution of the 16S gene or taxonomic groups with no evolutionary basis; the method does not attempt to signal the cause of such polyphyletic groups.
\citet{matsen2012reconciling} developed algorithms to quantify discordance between phylogeny and taxonomy based on a coloring problem previously described in the computer science literature \citep{moran2008convex}.
Although it is wonderful that several groups are actively working on taxonomic revision, it can be frustrating to have multiple different taxonomies with no easy way to translate between them or to the taxonomic names provided in the NCBI sequence database.

An obvious application of phylogenetics is to perform taxonomic classification, as the taxonomy is at least in part defined by phylogeny.
However, comparisons of taxonomic classification programs \citep{liu2008accurate,bazinet2012comparative} have indicated that current implementations of phylogenetic methods do not perform as well as simple classifiers based on counts of DNA substrings of a given length, which are often called $k$-mer classifiers \citep{wang2007naive,rosen2008metagenome}.
For those studies and others, the conventional metrics of classification performance such as precision and recall are applied to data simulated from known and taxonomically classified genomes.
\citet{srinivasan2012bacterial} find that phylogenetic methods to do taxonomic classification can outperform composition-based techniques at least for certain taxonomic groups.
% \citep{werner2011impact} Impact of training sets on classification of high-throughput bacterial 16s rRNA gene surveys
Some authors report that a combination of composition-based and homology-based classifiers work best \citep{brady2009phymm,parks2011classifying}.
The MEGAN program \citep{huson2007megan,huson2011integrative} BLASTs an unknown sequence onto a database of sequences with taxonomic labels and assigns the sequence the lowest (i.e.\ narrowest) taxonomic group shared by all of the high-quality hits; as such it is somewhat phylogenetic in that it uses the structure of the phylogenetic tree.
\citet{munch2008statistical,munch2008fast} infer taxonomic assignment by automatically retrieving sequences equipped with taxonomic information and building a tree on them along with an unknown sequence.
\citet{segata2012metagenomic} propose a clever approach to inferring organisms present in a metagenomic sample by compiling a database of clade-specific genes, then classifying a given read as being from the only clade that has the corresponding gene.
They show that this has good sensitivity and specificity, however, this method can only be used to identify the presence of organisms whose genome has been sequenced.
